NGO Repression as a Predictor of Worsening Human Rights Abuses

### 2021-11-15 manuscript/manuscript.Rmd
Expand Egypt case study

  canary <- tar_read(panel)
  tar_load(civicus_map_data)
  tar_load(civicus_clean)

@@

In exploring the impact of state crackdown on NGOs on subsequent human rights conditions, we look at both *de jure* measures—or formal legislation that seeks to repress NGOs—as well as the *de facto* implementation of these measures. This distinction is important—while states may pass anti-NGO laws, these laws can also lie dormant until governments feel the need to enforce them. For instance, Russia passed its 2015 Undesirable Organizations Law with the purpose of targeting and eliminating specific foreign-connected NGOs it deemed dangerous. One of the law's sponsors, Aleksandr Tarnavsky, described the law as a preventive measure that would not affect the majority of NGOs working in Russia. Rather, the law would be a "weapon hanging on the wall that never fires" and stand as a warning to potentially uncooperative NGOs [@Kozenko:2015]. Similarly, Egypt's 2002 NGO law instated severe barriers to freedom of association, banned NGOs from engaging in activities that "threaten national unity or violate public order or morality," and also prevented groups from receiving foreign funding. But despite the harsh legal framework, the Egyptian NGO sector nearly doubled in size between 1993–2011. This is because the state targeted a limited number of organizations working on politically sensitive issues such as human rights and democracy, while simultaneously tolerating organizations working on tamer issues [@Brechenmacher:2017].

@@

When holding all other factors equal, the de facto implementation of anti-civil society measures thus appear to be a stronger marginal indicator of future repression than the passage of formal de jure laws. In countries with low levels of political terror, it can potentially serve as a weak signal of a more general increase in political violence in the future, but predict more substantial changes in general respect for human rights. The difference in the effects of laws vs. their de facto implementation could also potentially be explained by the time frame of these strategies. De jure laws are a long-term strategy to prevent threats to the regime from coalescing in the first place. For instance, Egypt's Law 84 of 2002—which outlined many anti-NGO restrictions—has been on the books for nearly two decades, but only over the past five years have post-revolutionary regimes used it more broadly [@Ruffner:2015]. As such, the long-term effects of these laws in terms of their predictive power may require a much longer time period under study. In contrast, de facto crackdown on NGOs may signal the need to react sooner due to the immediacy of the threat.

@@

@@

Strengthening civil society is an important precondition for democratic transition and consolidation. This article shows dramatic growth in de jure regulations constraining NGOs; laws that can lay the foundation for subsequent de facto repression of civil society. Given these potential knock-on effects, in this paper, we posited that the ongoing global crackdown on NGOs could predict worsening human rights abuses. Using original data on de jure and de facto NGO repression from 1990–2013, we find weal support for the predictive power of de jure restriction on political terror and overall latent human rights values. Marginal predictions show that—all else held equal—adding new anti-NGO laws increases the probability of seeing moderate levels of political terror and is associated with worsening human rights conditions more generally. However, in contrast to de jure laws, de facto levels of NGO repression—including both legal and extra-legal means of crackdown—tend to increase the probability of more moderate political terror. Additionally, this repression is associated with worsening respect for broader human rights in the future.

@@

However, it is also important to state that while anti-NGO laws do not seem to be a predictor of broader human rights repression over the period under study, it does not imply they are not having a chilling effect on NGO operations. The broader literature on state repression of NGOs has shown that de jure laws can have numerous adverse consequences. Foreign funding laws have led to a 32% decline in bilateral aid flows [@DP:2018]. NGOs that are underfunded, understaffed tend to shut down when faced with increasing restrictions [@Heiss:2019]. In sum, while de jure laws do not effect levels of physical integrity and civil liberties violations, they have been shown to have other adverse consequences in the literature. Moreover, since the actual implementation of these laws through de facto civl society repression can have negative impacts of broader patterns of human rights abuses, the passage of these laws may still be a concerning first step that paves the path for implementation.

>>>>

  canary <- tar_read(panel)
  tar_load(vdem_clean)
  tar_load(civicus_map_data)
  tar_load(civicus_clean)

@@

In exploring the impact of state crackdown on NGOs on subsequent human rights conditions, we look at both *de jure* measures—or formal legislation that seeks to repress NGOs—as well as the *de facto* implementation of these measures. This distinction is important—while states may pass anti-NGO laws, these laws can also lie dormant until governments feel the need to enforce them. For instance, Russia passed its 2015 Undesirable Organizations Law with the purpose of targeting and eliminating specific foreign-connected NGOs it deemed dangerous. One of the law's sponsors, Aleksandr Tarnavsky, described the law as a preventive measure that would not affect the majority of NGOs working in Russia. Rather, the law would be a "weapon hanging on the wall that never fires" and stand as a warning to potentially uncooperative NGOs [@Kozenko:2015].

@@

When holding all other factors equal, the de facto implementation of anti-civil society measures thus appear to be a stronger marginal indicator of future repression than the passage of formal de jure laws. In countries with low levels of political terror, it can potentially serve as a weak signal of a more general increase in political violence in the future, but predict more substantial changes in general respect for human rights. The difference in the effects of laws vs. their de facto implementation could also potentially be explained by the time frame of these strategies. As we explore in the next section, de jure civil society laws are a long-term strategy to prevent threats to the regime from coalescing in the first place. As such, the long-term effects of these laws in terms of their predictive power may require a much longer time period under study. In contrast, de facto crackdown on NGOs may signal the need to react sooner due to the immediacy of the threat.

@@

# Discretionary disconnect between de jure and de facto NGO regulations: the case of Egypt

When looking both at marginal effects and improvements in predictive power, we thus find that the de facto implementation of anti-civil society measures are a stronger indicator of future repression than formal de jure laws. One reason for this disconnect between law and implementation is that laws targeted at NGOs are often not designed to be used immediately, but instead serve to protect regimes against potential threats in the future. Egypt provides a prime example of this mechanism.

```{r plot-egypt-cs, fig.width=6.5, fig.height=2, fig.asp=NULL, fig.cap="The disconnect between Egypt's de jure 2002 law and the widespread de facto repression of civil society a decade later"}
egypt_cs <- vdem_clean %>% 
  filter(gwcode == 651)

cs_color <- viridisLite::plasma(5, end = 0.8)[3]
law_color <- viridisLite::plasma(5, end = 0.8)[1]
event_color <- viridisLite::plasma(5, end = 0.8)[5]

ggplot(egypt_cs, aes(x = year, y = v2csreprss)) +
  geom_vline(xintercept = 2002, color = law_color) +
  geom_vline(xintercept = 2011, color = event_color) +
  geom_point(color = cs_color, size = 2) +
  geom_line(color = cs_color, size = 0.5) +
  annotate(geom = "segment", x = 2000, xend = 2002, y = -1.05, yend = -1.05, 
           arrow = arrow(angle = 15, length = unit(0.5, "lines"))) +
  annotate(geom = "segment", x = 2014, xend = 2011, y = -0.65, yend = -0.65, 
           arrow = arrow(angle = 15, length = unit(0.5, "lines"))) +
  annotate(geom = "label", x = 1999, y = -1.05, size = pts(9),
           label = "Law 84\nof 2002\npassed") +
  annotate(geom = "label", x = 2014, y = -0.65, size = pts(9),
           label = "2011\nuprising") +
  labs(x = "Year", y = "De facto CSO repression") +
  theme_ngo(9)
```

In the wake of domestic terrorist attacks in the late 1990s, the Egyptian parliament passed Law 84 of 2002, or the Law on Associations and Foundations [@Agati:2007, 60]. The new law imposed specific limits on international and foreign-connected NGOs and required that all NGOs register with the Ministry of Social Solidarity. Any revenue from foreign associations to nonprofit organizations working in Egypt required both recipients and donors to submit multiple applications to the Ministry of Foreign Affairs and state security, and the latter maintained ultimate veto power over the approval process [@Thabet:2004, 162]. Though it granted the government substantial oversight power over NGOs, Law 84 also allowed for enormous discretion in enforcement. In spite of the strict legal framework, the Egyptian NGO sector nearly doubled in size between 1993–2011 [@Brechenmacher:2017], largely because of the government's selective and targeted enforcement of the law. When NGOs become too contentious or pose a threat to the regime, officials could use minor infractions in more obscure parts of the law to fine or shut down the organization, but these attacks on NGOs were limited to specific NGOs. As seen in Figure \@ref(fig:plot-egypt-cs), Egypt saw its highest, most stable, and most open level of broader civil society repression in the decade following the passage of Law 84.

The calculus of enforcement changed as the state faced increased threats to its stability. Following Hosni Mubarak's removal as president in February 2011, the Supreme Council of the Armed Forces (SCAF) took temporary control over governing the country. SCAF worked throughout the year to curtail protests and often resorted to violence to break up demonstrations, which then led to larger and more violent protests in response. The tension culminated on December 16, when troops filled Tahrir Square to put down an anti-SCAF demonstration, killing 10 and wounding more than 400. Two weeks later, Egyptian security forces raided the offices of 17 local and international NGOs (Fadel and Warrick 2011)—including Freedom House, the National Democratic Institute, and the International Republican Institute—and arrested dozens of domestic and international NGO staff. The stated legal pretext for this raid was that the organizations were not officially registered and thus in violation of Law 84. The de jure law—constantly looming in the background, but enforced selectively—was finally implemented and enforced more expansively. In the years after the 2011 uprising, post-revolutionary regimes have enforced the law more consistently, further constricting the space for civil society [@Ruffner:2015], as seen in the rapid drop in broader civil society repression in Figure \@ref(fig:plot-egypt-cs).

Law 84 (and its replacement Law 70 of 2017) is thus a useful lever for selectively controlling the allowable space for dissidence and leaving open the possibility of future human rights abuses. The law's built-in discretionary authority permits the government to give the appearance of concessions and openness to civil society while simultaneously providing the government with the ability to target individual NGOs and reserving the right to engage in broader crackdowns of associational rights. This trend holds across many other countries—anti-NGO laws can serve as a useful avenue for future human rights abuses, sometimes with years separating the official passage of the law and the large-scale implementation of the law.

@@

Strengthening civil society is an important precondition for democratic transition and consolidation. This article shows dramatic growth in de jure regulations constraining NGOs; laws that can lay the foundation for subsequent de facto repression of civil society. Given these potential knock-on effects, in this paper, we posited that the ongoing global crackdown on NGOs could predict worsening human rights abuses. Using original data on de jure and de facto NGO repression from 1990–2013, we find weak support for the predictive power of de jure restriction on political terror and overall latent human rights values. Marginal predictions show that—all else held equal—adding new anti-NGO laws increases the probability of seeing moderate levels of political terror and is associated with worsening human rights conditions more generally. However, in contrast to de jure laws, de facto levels of NGO repression—including both legal and extra-legal means of crackdown—tend to increase the probability of more moderate political terror. Additionally, this repression is associated with worsening respect for broader human rights in the future.

@@

However, it is also important to state that while anti-NGO laws do not seem to be a predictor of broader human rights repression over the period under study, it does not imply they are not having a chilling effect on NGO operations. The broader literature on state repression of NGOs has shown that de jure laws can have numerous adverse consequences. Foreign funding laws have led to a 32% decline in bilateral aid flows [@DP:2018]. NGOs that are underfunded, understaffed tend to shut down when faced with increasing restrictions [@Heiss:2019]. In sum, while de jure laws do not effect levels of physical integrity and civil liberties violations in particular, research has found that they other adverse consequences. Moreover, since the actual implementation of these laws through de facto civl society repression can have negative effects on broader patterns of human rights abuses, the passage of these laws may still be a concerning first step that paves the path for future implementation.
